---
title: "Calgary Flood Inundation Probability Modeling Workflow"
format:
  html:
    toc: true
    toc-depth: 3
    number-sections: true
execute:
  echo: true
  warning: false
  message: false
---

# Setup / libraries / reproducibility notes

This report builds a Calgary-only flood inundation probability workflow in R. It creates a 30 m fishnet, derives the binary inundation target, engineers all required features, performs exploratory checks and minor cleaning, fits a `glm` logistic regression model, and saves validation outputs for later reuse.

```{r}
# Locate the project root, register the project-local R library, and load packages.
report_dir <- normalizePath(dirname(knitr::current_input(dir = TRUE)), winslash = "/", mustWork = TRUE)
project_root <- normalizePath(file.path(report_dir, ".."), winslash = "/", mustWork = TRUE)
.libPaths(c(normalizePath(file.path(project_root, "r_libs")), .libPaths()))

# Set chunk defaults so the report is readable and reproducible.
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.width = 8,
  fig.height = 6,
  dpi = 150
)

# Load the packages used across the workflow.
required_packages <- c(
  "sf", "terra", "exactextractr", "whitebox", "dplyr", "tidyr", "readr",
  "purrr", "stringr", "ggplot2", "broom", "pROC", "car", "janitor",
  "scales", "forcats", "knitr"
)
invisible(lapply(required_packages, library, character.only = TRUE))
sf::sf_use_s2(FALSE)

# Source the helper scripts that own each major stage of the workflow.
source(file.path(project_root, "R", "01_setup_target.R"))
source(file.path(project_root, "R", "02_features_topography_hydrology_watershed.R"))
source(file.path(project_root, "R", "03_features_landcover_built.R"))
source(file.path(project_root, "R", "04_merge_eda_clean.R"))
source(file.path(project_root, "R", "05_model_diagnostics.R"))

# Read the boundary once so the report can lock the projected analysis CRS.
boundary_meta <- st_read(file.path(project_root, "data", "calgary_boundary.shp"), quiet = TRUE)
analysis_crs_input <- st_crs(boundary_meta)$input

# Define the stable predictor names that later City X work will reuse.
predictor_cols <- c(
  "mean_elev", "min_elev", "elev_range", "mean_slope", "max_slope", "sd_elev",
  "dist_nearest_stream", "water_cover_area", "river_density",
  "max_flow_accum", "upstream_ws_area",
  "impervious_ratio", "vegetation_ratio", "open_soil_ratio", "agriculture_ratio",
  "building_cover_ratio"
)

# Build one shared config list, then override thresholds by stage where needed.
base_config <- list(
  project_root = project_root,
  fishnet_size_m = 30,
  seed = 20250322L,
  target_col = "target",
  predictor_cols = predictor_cols,
  analysis_crs_input = analysis_crs_input,
  landcover_groups = list(
    vegetation = c("Forest", "Grassland", "Shrubland", "Manicured", "GolfCourse", "Natural Wetland"),
    open_soil = c("Bare Ground"),
    agriculture = c("Agricultural_Annual Crops", "Agricultural_Pasture/Fallow")
  ),
  impervious_values = c("Concrete", "Pavement"),
  paths = list(
    data_dir = file.path(project_root, "data"),
    intermediate_dir = file.path(project_root, "outputs", "intermediate"),
    models_dir = file.path(project_root, "outputs", "models"),
    diagnostics_dir = file.path(project_root, "outputs", "diagnostics"),
    figures_dir = file.path(project_root, "outputs", "figures"),
    eda_fig_dir = file.path(project_root, "outputs", "figures", "eda"),
    diag_fig_dir = file.path(project_root, "outputs", "figures", "diagnostics"),
    fishnet_base = file.path(project_root, "outputs", "intermediate", "fishnet_30m.gpkg"),
    fishnet_target = file.path(project_root, "outputs", "intermediate", "fishnet_target.gpkg"),
    target_summary = file.path(project_root, "outputs", "intermediate", "fishnet_target_summary.csv"),
    features_topography = file.path(project_root, "outputs", "intermediate", "features_topography.csv"),
    features_hydrology = file.path(project_root, "outputs", "intermediate", "features_hydrology.csv"),
    features_watershed = file.path(project_root, "outputs", "intermediate", "features_watershed.csv"),
    features_landcover = file.path(project_root, "outputs", "intermediate", "features_landcover.csv"),
    features_built = file.path(project_root, "outputs", "intermediate", "features_built.csv"),
    features_all = file.path(project_root, "outputs", "intermediate", "features_all.gpkg"),
    modeling_raw = file.path(project_root, "outputs", "intermediate", "modeling_table_raw.csv"),
    modeling_clean = file.path(project_root, "outputs", "intermediate", "modeling_table_clean.csv"),
    model_rds = file.path(project_root, "outputs", "models", "glm_logistic_calgary.rds"),
    scaling_params = file.path(project_root, "outputs", "diagnostics", "scaling_parameters.csv"),
    model_summary = file.path(project_root, "outputs", "diagnostics", "model_summary.csv"),
    confusion_metrics = file.path(project_root, "outputs", "diagnostics", "confusion_metrics.csv"),
    outcome_definitions = file.path(project_root, "outputs", "diagnostics", "tp_tn_fp_fn_definitions.csv"),
    roc_data = file.path(project_root, "outputs", "diagnostics", "roc_curve_data.csv"),
    roc_plot = file.path(project_root, "outputs", "figures", "diagnostics", "roc_curve.png"),
    predictions_full = file.path(project_root, "outputs", "intermediate", "predictions_full.csv"),
    train_eval = file.path(project_root, "outputs", "intermediate", "train_eval.csv"),
    test_eval = file.path(project_root, "outputs", "intermediate", "test_eval.csv"),
    prediction_map = file.path(project_root, "outputs", "figures", "diagnostics", "prediction_map.png"),
    train_outcome_map = file.path(project_root, "outputs", "figures", "diagnostics", "train_outcome_map.png"),
    test_outcome_map = file.path(project_root, "outputs", "figures", "diagnostics", "test_outcome_map.png"),
    missingness_csv = file.path(project_root, "outputs", "diagnostics", "missingness_summary.csv"),
    corr_csv = file.path(project_root, "outputs", "diagnostics", "correlation_matrix.csv"),
    corr_png = file.path(project_root, "outputs", "figures", "eda", "correlation_matrix.png"),
    variance_csv = file.path(project_root, "outputs", "diagnostics", "variance_check.csv"),
    outlier_csv = file.path(project_root, "outputs", "diagnostics", "outlier_summary.csv"),
    class_balance_csv = file.path(project_root, "outputs", "diagnostics", "class_balance.csv"),
    target_map_png = file.path(project_root, "outputs", "figures", "eda", "target_map.png"),
    target_valid_map_png = file.path(project_root, "outputs", "figures", "eda", "target_share_map.png"),
    whitebox_dir = file.path(project_root, "outputs", "intermediate", "whitebox")
  )
)

target_config <- base_config
target_config$classification_threshold <- 0.30

model_config <- base_config
model_config$classification_threshold <- 0.50
```

# Data inputs

The workflow uses only the canonical Calgary inputs from the top-level `data/` folder. The helper functions print their own QA messages, but this chunk records the file inventory and the observed raster coding used by the report.

```{r}
# Build a compact input inventory so the report documents the actual source files.
input_inventory <- tibble::tibble(
  file = c(
    "calgary_boundary.shp",
    "inundation_calgary.tif",
    "calgary_dem.tif",
    "calgary_waterways.shp",
    "calgary_surf_water.tif",
    "calgary_Imp_surf.geojson",
    "calgary_landcover.geojson",
    "calgary_building.shp"
  ),
  path = file.path(base_config$paths$data_dir, file),
  exists = file.exists(path),
  size_mb = round(file.info(path)$size / 1024^2, 2)
)
knitr::kable(input_inventory, caption = "Canonical Calgary inputs used in this workflow")

# Inspect the target and surface-water raster coding before any modeling work begins.
inundation_rast <- terra::rast(file.path(base_config$paths$data_dir, "inundation_calgary.tif"))
surface_water_rast <- terra::rast(file.path(base_config$paths$data_dir, "calgary_surf_water.tif"))

inundation_values <- terra::freq(inundation_rast, digits = 0)
surface_water_values <- terra::freq(surface_water_rast, digits = 0)

knitr::kable(inundation_values, caption = "Observed inundation raster values")
knitr::kable(surface_water_values, caption = "Observed surface-water raster values")
```

The inundation raster behaves as a binary target layer with `0` and `1` as valid values and a separate excluded code. The surface-water raster is also binary, so the hydrology step can compute `water_cover_area` without further class-mapping assumptions.

# Fishnet creation

```{r}
# Create the 30 m Calgary fishnet aligned to the boundary CRS and write the base output.
fishnet_results <- build_fishnet_and_target(target_config)
fishnet_base <- st_read(base_config$paths$fishnet_base, quiet = TRUE)

# Print a few checks so the report shows the grid structure explicitly.
fishnet_base_summary <- tibble::tibble(
  n_rows = nrow(fishnet_base),
  unique_ids = dplyr::n_distinct(fishnet_base$id),
  crs = st_crs(fishnet_base)$input
)
knitr::kable(fishnet_base_summary, caption = "Base 30 m fishnet checks")
```

The base fishnet contains one row per Calgary grid cell in the projected local CRS. Later steps only keep cells with valid inundation coverage, but the base grid is still saved for reuse and inspection.

# Target variable generation

```{r}
# Read the target-valid fishnet and the saved QA summary from the target-generation step.
fishnet_target <- st_read(base_config$paths$fishnet_target, quiet = TRUE)
target_summary <- readr::read_csv(base_config$paths$target_summary, show_col_types = FALSE)
knitr::kable(target_summary, caption = "Target-generation summary for Calgary")

# Plot the observed target so the report shows the modeled response variable.
ggplot(fishnet_target) +
  geom_sf(aes(fill = factor(target)), color = NA) +
  scale_fill_manual(values = c("0" = "#2b8cbe", "1" = "#de2d26"), name = "Target") +
  labs(title = "Observed binary inundation target in Calgary") +
  theme_void(base_size = 11) +
  theme(legend.position = "right")
```

The target layer keeps only fishnet cells with valid raster coverage and converts inundation share to a binary outcome using the required `0.30` cutoff. That step removes the unsupported northern cells before any feature engineering begins.

# Feature engineering

## Topography

```{r}
# Run the DEM, hydrology, and watershed feature engineering helper.
feature_results <- compute_topography_hydrology_watershed(base_config)

topography_tbl <- readr::read_csv(base_config$paths$features_topography, show_col_types = FALSE)
knitr::kable(dplyr::slice_head(topography_tbl, n = 6), caption = "Topography feature sample")
```

The topography block derives elevation and slope summaries directly from the Calgary DEM. Slope is kept in degrees throughout the workflow so the units remain consistent between feature engineering and interpretation.

## Hydrology

```{r}
# Read the saved hydrology outputs so they can be inspected without recomputing them in later chunks.
hydrology_tbl <- readr::read_csv(base_config$paths$features_hydrology, show_col_types = FALSE)
knitr::kable(dplyr::slice_head(hydrology_tbl, n = 6), caption = "Hydrology feature sample")
```

These hydrology variables combine stream proximity, surface-water coverage, and within-cell stream density. They are all computed in projected meter units, which keeps distances, lengths, and densities interpretable.

## Watershed

```{r}
# Read the watershed output table produced by the hydrology helper.
watershed_tbl <- readr::read_csv(base_config$paths$features_watershed, show_col_types = FALSE)
knitr::kable(dplyr::slice_head(watershed_tbl, n = 6), caption = "Watershed feature sample")
```

The watershed block is the most computationally demanding part of the workflow. It caches the square-grid hydrology intermediates so later reruns can reuse the filled DEM, flow-direction, flow-accumulation, and watershed products.

## Land cover

```{r}
# Run the land-cover and built-environment helper using the conservative class mappings.
landcover_results <- compute_landcover_built_features(base_config)
landcover_tbl <- readr::read_csv(base_config$paths$features_landcover, show_col_types = FALSE)
knitr::kable(dplyr::slice_head(landcover_tbl, n = 6), caption = "Land-cover ratio sample")
```

The land-cover ratios use the conservative mapping agreed earlier: vegetation includes forest, grassland, shrubland, manicured land, golf course, and natural wetland; open soil is bare ground only; agriculture is the two agricultural classes.

## Built environment

```{r}
# Read the built-environment output that combines impervious and building coverage ratios.
built_tbl <- readr::read_csv(base_config$paths$features_built, show_col_types = FALSE)
knitr::kable(dplyr::slice_head(built_tbl, n = 6), caption = "Built-environment ratio sample")
```

Impervious coverage is restricted to `Concrete` and `Pavement`, while building coverage comes directly from the building-footprint layer. Both ratios are saved with the same `id` key used by every other stage.

# Merge final fishnet feature table

```{r}
# Merge all feature tables back into the target-valid fishnet and save the raw modeling table.
merge_results <- merge_and_clean_features(base_config)
fishnet_features <- st_read(base_config$paths$features_all, quiet = TRUE)
modeling_raw <- readr::read_csv(base_config$paths$modeling_raw, show_col_types = FALSE)

merged_summary <- tibble::tibble(
  n_feature_rows = nrow(fishnet_features),
  n_modeling_rows = nrow(modeling_raw),
  unique_ids = dplyr::n_distinct(modeling_raw$id),
  n_predictors = length(setdiff(names(modeling_raw), c("id", "target", "inundated_share")))
)
knitr::kable(merged_summary, caption = "Merged feature table checks")
```

At this stage the workflow has one geospatial feature layer for mapping and one flat table for modeling. Both share the same stable field names so the same pipeline can later be reused for City X.

# EDA

```{r}
# Read the EDA summary tables written by the merge-and-clean helper.
missingness_tbl <- readr::read_csv(base_config$paths$missingness_csv, show_col_types = FALSE)
variance_tbl <- readr::read_csv(base_config$paths$variance_csv, show_col_types = FALSE)
outlier_tbl <- readr::read_csv(base_config$paths$outlier_csv, show_col_types = FALSE)
class_balance_tbl <- readr::read_csv(base_config$paths$class_balance_csv, show_col_types = FALSE)

knitr::kable(missingness_tbl, caption = "Missingness summary")
knitr::kable(class_balance_tbl, caption = "Target class balance")
knitr::kable(variance_tbl, caption = "Zero and near-zero variance check")
knitr::kable(dplyr::slice_head(outlier_tbl, n = 10), caption = "Outlier screening summary")
```

The EDA step is intentionally conservative. It documents missingness, class balance, variance problems, and extreme values without over-cleaning the dataset or changing the feature definitions in a way that would make later transfer difficult.

```{r}
# Display the saved correlation heatmap and selected spatial figures when they are available.
if (file.exists(base_config$paths$corr_png)) {
  knitr::include_graphics(base_config$paths$corr_png)
}

selected_maps <- c(
  base_config$paths$target_map_png,
  file.path(base_config$paths$eda_fig_dir, "map_mean_elev.png"),
  file.path(base_config$paths$eda_fig_dir, "map_upstream_ws_area.png"),
  file.path(base_config$paths$eda_fig_dir, "map_impervious_ratio.png")
)
selected_maps <- selected_maps[file.exists(selected_maps)]
if (length(selected_maps) > 0) {
  knitr::include_graphics(selected_maps)
}
```

These figures highlight both the response variable and a representative set of predictors. The selected feature maps include one watershed variable, which is important for the later modeling discussion even if that term is not among the strongest coefficients.

# Data cleaning

```{r}
# Read the cleaned modeling table and report the retained predictors after minor cleaning.
modeling_clean <- readr::read_csv(base_config$paths$modeling_clean, show_col_types = FALSE)
cleaning_summary <- tibble::tibble(
  raw_rows = nrow(modeling_raw),
  cleaned_rows = nrow(modeling_clean),
  cleaned_predictors = length(setdiff(names(modeling_clean), c("id", "target", "inundated_share")))
)
knitr::kable(cleaning_summary, caption = "Minor-cleaning summary")
```

Cleaning removes invalid target rows, missing predictor rows, and zero-variance fields if they appear. The workflow avoids aggressive transformations so the same feature set remains portable to later prediction tasks.

# Logistic regression modeling

```{r}
# Fit the Calgary glm workflow using training-only scaling and a fixed 0.5 classification threshold.
model_results <- run_glm_diagnostics(model_config)
model_summary_tbl <- readr::read_csv(base_config$paths$model_summary, show_col_types = FALSE)
knitr::kable(model_summary_tbl, caption = "Final logistic regression model summary")
```

The model stage standardizes predictors using the training subset only, then fits a logistic regression with documented VIF and AIC screening. The `0.5` classification threshold is used only for validation outputs; the target itself was still created with the required `0.30` inundation-share rule.

# Validation and diagnostics

```{r}
# Read the saved validation outputs so the report can present them in table form.
confusion_tbl <- readr::read_csv(base_config$paths$confusion_metrics, show_col_types = FALSE)
outcome_definitions_tbl <- readr::read_csv(base_config$paths$outcome_definitions, show_col_types = FALSE)
knitr::kable(confusion_tbl, caption = "Confusion metrics on the test split")
knitr::kable(outcome_definitions_tbl, caption = "Plain-language outcome definitions")
```

The validation stage reports both model discrimination and classification performance. The confusion table makes it easier to see the operational tradeoff between missed flooded cells and false flood alarms.

```{r}
# Include the ROC curve and the main diagnostic maps saved by the modeling helper.
diagnostic_figures <- c(
  base_config$paths$roc_plot,
  base_config$paths$prediction_map,
  base_config$paths$train_outcome_map,
  base_config$paths$test_outcome_map
)
diagnostic_figures <- diagnostic_figures[file.exists(diagnostic_figures)]
if (length(diagnostic_figures) > 0) {
  knitr::include_graphics(diagnostic_figures)
}
```

The probability map shows how the fitted model generalizes across Calgary, while the outcome maps break predictions into TP, TN, FP, and FN so spatial error patterns are easy to inspect.

# Key findings / discussion

The Calgary workflow is fully modular: every stage writes an intermediate artifact keyed by `id`, and the final modeling table preserves the same predictor names that will be needed later for City X transfer. The workflow also makes all major assumptions explicit, including the conservative land-cover mapping, the binary raster handling rules, and the separate thresholds used for target creation versus classification.

The strongest practical value of this report is transparency. Fishnet creation, target generation, feature engineering, EDA, cleaning, modeling, and diagnostics are all separated into named helper stages, which makes reruns and future debugging much easier than a single monolithic script.

# Limitations and next steps

This Calgary-only workflow still has two important limitations. First, the watershed stage is computationally heavy because it relies on centroid-based hydrologic routing at city scale. Second, the final model quality should be interpreted with the usual caution for transfer learning: even a well-performing Calgary model may require recalibration before it is applied in another city.

The next step for City X deployment is to reproduce the same predictor set with the same variable names and the same preprocessing contract. Once those City X features exist, the saved scaling rules and `glm` object can be used to produce probability surfaces without changing the Calgary training workflow.
